High-level cognition during story listening is reflected in high-order dynamic correlations in neural activity patterns

### 2021-08-02 paper/main.tex
WIP PR comments

\subsection*{Recovering dynamic first-order correlations}
We generated synthetic datasets that differed in how the underlying first-order
correlations changed over time.  For each dataset, we applied
Equation~\ref{eqn:timecorr} with a variety of kernel shapes and
widths.  We assessed how well the true underlying correlations at each
timepoint matched the recovered correlations
(Fig.~\ref{fig:synthetic}).  For every kernel and dataset we tested,
our approach recovered the correlation dynamics we embedded into the
data.  However, the quality of these recoveries varied across
different synthetic datasets in a kernel-dependent way.

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/synthetic_data}
  \caption{\textbf{Recovering known dynamic first-order correlations
       from synthetic data.} Each panel displays the average
    correlations between the vectorized upper triangles of the
    recovered correlation matrix at each timepoint and either the true
    underlying correlation at each timepoint or a reference
    correlation matrix.  (The averages are taken across 100 different
    randomly generated synthetic datasets of the given category.)
    Error ribbons denote 95\% confidence intervals of the mean (taken across
    datasets). Different colors denote different kernel shapes, and
    the shading within each color family denotes the kernel width
    parameter.  For a complete description of each synthetic dataset,
    see Synthetic data: simulating dynamic first-order
      correlations. \textbf{a. Constant correlations.}  These
    datasets have a stable (unchanging) underlying correlation matrix.
  \textbf{b. Random correlations.}  These datasets are generated
    using a new independently drawn correlation matrix at each new
    timepoint. \textbf{c. Ramping correlations.} These datasets are
    generated by smoothly varying the underlying correlations between
    the randomly drawn correlation matrices at the first and last
    timepoints.  The left panel displays the correlations between the
    recovered dynamic correlations and the underlying ground truth
    correlations.  The middle panel compares the recovered
    correlations with the first timepoint's correlation
    matrix.  The right panel compares the recovered correlations with
    the last timepoint's correlation matrix.
  \textbf{d. Event-based correlations.} These datasets are each
    generated using five randomly drawn correlation matrices that each
    remain stable for a fifth of the total timecourse.  The left panel
    displays the correlations between the recovered dynamic
    correlations and the underlying ground truth correlations.  The
    right panels compare the recovered correlations with the
    correlation matrices unique to each event.  The vertical lines
    denote event boundaries. Source data are provided as a Source Data file.}
  \label{fig:synthetic}
\end{figure}

@@

\subsection*{Recovering dynamic higher-order correlations}
Following our approach to evaluating our ability to recover known
dynamic first-order correlations from synthetic data, we generated an
analogous second set of synthetic datasets that we designed to exhibit
known dynamic first-order and second-order correlations (see
Synthetic data: simulating dynamic higher-order
  correlations).  We generated a total of 400 datasets (100 datasets
for each category) that varied in
how the first-order and second-order correlations changed over time.
We then repeatedly applied Equation~\ref{eqn:timecorr} using the
overall best-performing kernel from our first-order tests (a Laplace
kernel with a width of 20; Fig.~\ref{fig:synthetic}) to assess how
closely the recovered dynamic correlations matched the dynamic
correlations we had embedded into the datasets.

Overall, we found that we could reliably recover both first-order and
second-order correlations from the synthetic data
(Fig.~\ref{fig:higher_order_sims}).  When the correlations were stable
for longer intervals, or changed gradually (constant, ramping, and
event datasets), recovery performance was relatively high, and we were
better able to recover dynamic first-order correlations than
second-order correlations.  This is because errors in our estimation
procedure at lower orders necessarily propagate to higher orders
(since lower-order correlations are used to estimate higher-order
correlations).  Conversely, when the correlations were particularly
unstable (random datasets), we better recovered second-order
correlations.  This is because noises in our data generation
procedure propagates from higher orders to lower orders (see
Synthetic data: simulating dynamic high-order correlations).

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/higher_order_sims}
  \caption{\textbf{Recovery of simulated first-order and second-order dynamic correlations.} Each panel displays the average correlations
    between the vectorized upper triangles of the recovered
    first-order and second-order correlation matrices and the true
    (simulated) first-order and second order correlation matrices at each
    timepoint and for each synthetic dataset.  (The averages are taken across 100 different
    randomly generated synthetic datasets of the given category.)
    Error ribbons denote 95\% confidence intervals of the mean (taken across
    datasets).  For a complete description of each synthetic dataset,
    see Synthetic data: simulating dynamic higher-order
      correlations.  All estimates represented in this figure were
    computed using a Laplace kernel
    (width~$=20$).  \textbf{Constant.} These datasets
    have stable (unchanging) underlying second-order correlation
    matrices. \textbf{Random.} These datasets are
    generated using a new independently drawn second-order correlation
    matrix at each timepoint. \textbf{Ramping.}
    These datasets are generated by smoothly varying the underlying
    second-order correlations between the randomly drawn correlation
    matrices at the first and last timepoints. \textbf{Event.}  These datasets are each generated using five
    randomly drawn second-order correlation matrices that each remain
    stable for a fifth of the total timecourse.  The
    vertical lines denote event boundaries.  Note that the ``dips''
    and ``ramps'' at the boundaries of sharp transitions (e.g., the
    beginning and ends of the ``constant'' and ``ramping'' datasets,
    and at the event boundaries of the ``event'' datasets) are
    finite-sample effects that reflect the reduced numbers of samples
    that may be used to accurately estimate correlations at sharp
    boundaries. Source data are provided as a Source Data file.}
  \label{fig:higher_order_sims}
\end{figure}

@@

\subsection*{Cognitively relevant dynamic high-order correlations in
  fMRI data}
We used across-participant temporal decoders to identify cognitively
relevant neural patterns in fMRI data (see Forward inference
  and decoding accuracy).  The dataset we examined~\citep[collected
by][]{SimoEtal16} comprised four experimental conditions that exposed
participants to stimuli that varied systematically in how cognitively
engaging they were.  The intact experimental condition had
participants listen to an audio recording of a 10-minute story.  The
paragraph-scrambled experimental condition had participants
listen to a temporally scrambled version of the story, where the
paragraphs occurred out of order (but where the same total set of
paragraphs were presented over the full listening interval).  All
participants in this condition experienced the scrambled paragraphs in
the same order.  The word-scrambled experimental condition
had participants listen to a temporally scrambled version of the story
where the words in the story occurred in a random order.  All
participants in the word condition experienced the scrambled words in
the same order.  Finally, in a rest experimental condition,
participants lay in the scanner with no overt stimulus, with their
eyes open (blinking as needed).  This public dataset provided a convenient
means of testing our hypothesis that different levels of cognitive
processing and engagement are reflected in different orders of brain
activity dynamics.

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/decode_level}
  \caption{\textbf{Across-participant timepoint decoding accuracy varies with
       correlation order and cognitive engagement.}
     \textbf{a.~Decoding accuracy as a function of order: PCA.}
    Order ($x$-axis) refers to the maximum order of dynamic
    correlations that were available to the classifiers (see
    Feature weighting and testing).  The reported
    across-participant decoding accuracies are averaged over all
    kernel shapes and widths (see Identifying robust decoding
      results).  The $y$-values are displayed relative to chance
    accuracy (intact: $\frac{1}{300}$; paragraph: $\frac{1}{272}$;
    word: $\frac{1}{300}$; rest: $\frac{1}{400}$; these chance accuracies
    were subtracted from the observed accuracies to obtain the
    relative accuracies reported on the $y$-axis).  The error ribbons
    denote 95\% confidence intervals of the means across cross-validation folds
    (i.e., random assignments of participants to the training and test
    sets).  The colors denote the experimental condition.  Arrows
    denote sets of features that yielded reliably higher (upward
    facing) or lower (downward facing) decoding accuracy than the mean
    of all other features (via a two-tailed $t$-test, thresholded at
    $p < 0.05$).  Figure~\ref{fig:ttests} displays additional
    comparisons between the decoding accuracies achieved using
    different sets of neural features.  The circled values represent
    the maximum decoding accuracy within each experimental condition.
  \textbf{b.~Normalized timepoint decoding accuracy as a function of order:
       PCA.}  This panel displays the same results as Panel a, but here
    each curve has been normalized to have a maximum value of 1 and a
    minimum value of 0 (including the upper and lower bounds of the
    respective 95\% confidence intervals of the mean).  Panels a and b used PCA to
    project each high-dimensional pattern of dynamic correlations onto
    a lower-dimensional space. \textbf{c.~Timepoint decoding accuracy as a
       function of order: eigenvector centrality.} This panel is in the
    same format as Panel a, but here eigenvector centrality has been
    used to project the high-dimensional patterns of dynamic
    correlations onto a lower-dimensional space.
 \textbf{d.~Normalized timepoint decoding accuracy as a function of order:
       eigenvector centrality.}  This panel is in the same format as
    Panel b, but here eigenvector centrality has been used to project
    the high-dimensional patterns of dynamic correlations onto a
    lower-dimensional space. See
    Figures~\kernelshape~and~\kernelwidth~for decoding results broken
    down by kernel shape and width, respectively. Source data are provided as a Source Data file.}
  \label{fig:decoding}
\end{figure}

>>>>

\subsection*{Recovering known dynamic first-order correlations}
We generated synthetic datasets that differed in how the underlying first-order
correlations changed over time.  For each dataset, we applied
Equation~\ref{eqn:timecorr} with a variety of kernel shapes and
widths.  We assessed how well the true underlying correlations at each
timepoint matched the recovered correlations
(Fig.~\ref{fig:synthetic}).  For every kernel and dataset we tested,
our approach recovered the correlation dynamics we embedded into the
data.  However, the quality of these recoveries varied across
different synthetic datasets in a kernel-dependent way.

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/synthetic_data}
  \caption{\textbf{Recovering known dynamic first-order correlations
       from synthetic data.} Each panel displays the average
    correlations between the vectorized upper triangles of the
    recovered correlation matrix at each timepoint and either the true
    underlying correlation at each timepoint or a reference
    correlation matrix.  (The averages are taken across 100 different
    randomly generated synthetic datasets of the given category, each
    with $K = 50$ features and $T = 300$ timepoints.)
    Error ribbons denote 95\% confidence intervals of the mean (taken across
    datasets). Different colors denote different kernel shapes, and
    the shading within each color family denotes the kernel width
    parameter.  For a complete description of each synthetic dataset,
    see Synthetic data: simulating dynamic first-order
      correlations. \textbf{a. Constant correlations.}  These
    datasets have a stable (unchanging) underlying correlation matrix.
  \textbf{b. Random correlations.}  These datasets are generated
    using a new independently drawn correlation matrix at each new
    timepoint. \textbf{c. Ramping correlations.} These datasets are
    generated by smoothly varying the underlying correlations between
    the randomly drawn correlation matrices at the first and last
    timepoints.  The left panel displays the correlations between the
    recovered dynamic correlations and the underlying ground truth
    correlations.  The middle panel compares the recovered
    correlations with the first timepoint's correlation
    matrix.  The right panel compares the recovered correlations with
    the last timepoint's correlation matrix.
  \textbf{d. Event-based correlations.} These datasets are each
    generated using five randomly drawn correlation matrices that each
    remain stable for a fifth of the total timecourse.  The left panel
    displays the correlations between the recovered dynamic
    correlations and the underlying ground truth correlations.  The
    right panels compare the recovered correlations with the
    correlation matrices unique to each event.  The vertical lines
    denote event boundaries. Source data are provided as a Source Data file.}
  \label{fig:synthetic}
\end{figure}

@@

\subsection*{Recovering known dynamic higher-order correlations}
Following our approach to evaluating our ability to recover known
dynamic first-order correlations from synthetic data, we generated an
analogous second set of synthetic datasets that we designed to exhibit
known dynamic first-order and second-order correlations (see
Synthetic data: simulating dynamic higher-order
  correlations).  We generated a total of 400 datasets (100 datasets
for each category) that varied in
how the first-order and second-order correlations changed over time.
We then repeatedly applied Equation~\ref{eqn:timecorr} using the
overall best-performing kernel from our first-order tests (a Laplace
kernel with a width of 20; Fig.~\ref{fig:synthetic}) to assess how
closely the recovered dynamic correlations matched the dynamic
correlations we had embedded into the datasets.

Overall, we found that we could reliably recover both first-order and
second-order correlations from the synthetic data
(Fig.~\ref{fig:higher_order_sims}).  When the correlations were stable
for longer intervals, or changed gradually (constant, ramping, and
event datasets), recovery performance was relatively high, and we were
better able to recover dynamic first-order correlations than
second-order correlations.  This is because errors in our estimation
procedure at lower orders necessarily propagate to higher orders
(since lower-order correlations are used to estimate higher-order
correlations).  Conversely, when the correlations were particularly
unstable (random datasets), we better recovered second-order
correlations.  This is because noise in our data generation
procedure propagates from higher orders to lower orders (see
Synthetic data: simulating dynamic high-order correlations).

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/higher_order_sims}
  \caption{\textbf{Recovery of simulated first-order and second-order dynamic correlations.} Each panel displays the average correlations
    between the vectorized upper triangles of the recovered
    first-order and second-order correlation matrices and the true
    (simulated) first-order and second order correlation matrices at each
    timepoint and for each synthetic dataset.  (The averages are taken across 100 different
    randomly generated synthetic datasets of the given category, each
    with $K = 10$ features and $T = 300$ timepoints.)
    Error ribbons denote 95\% confidence intervals of the mean (taken across
    datasets).  For a complete description of each synthetic dataset,
    see Synthetic data: simulating dynamic higher-order
      correlations.  All estimates represented in this figure were
    computed using a Laplace kernel
    (width~$=20$).  \textbf{Constant.} These datasets
    have stable (unchanging) underlying second-order correlation
    matrices. \textbf{Random.} These datasets are
    generated using a new independently drawn second-order correlation
    matrix at each timepoint. \textbf{Ramping.}
    These datasets are generated by smoothly varying the underlying
    second-order correlations between the randomly drawn correlation
    matrices at the first and last timepoints. \textbf{Event.}  These datasets are each generated using five
    randomly drawn second-order correlation matrices that each remain
    stable for a fifth of the total timecourse.  The
    vertical lines denote event boundaries.  Note that the ``dips''
    and ``ramps'' at the boundaries of sharp transitions (e.g., the
    beginning and ends of the ``constant'' and ``ramping'' datasets,
    and at the event boundaries of the ``event'' datasets) are
    finite-sample effects that reflect the reduced numbers of samples
    that may be used to accurately estimate correlations at sharp
    boundaries. Source data are provided as a Source Data file.}
  \label{fig:higher_order_sims}
\end{figure}

@@

\subsection*{Cognitively relevant dynamic high-order correlations in
  fMRI data}
We used across-participant temporal decoders to identify cognitively
relevant neural patterns in fMRI data (see Forward inference
  and decoding accuracy).  The dataset we examined~\citep[collected
by][]{SimoEtal16} comprised four experimental conditions that exposed
participants to stimuli that varied systematically in how cognitively
engaging they were.  The intact experimental condition (intact) had
participants listen to an audio recording of a 10-minute story.  The
paragraph-scrambled experimental condition (paragraph) had participants
listen to a temporally scrambled version of the story, where the
paragraphs occurred out of order (but where the same total set of
paragraphs were presented over the full listening interval).  All
participants in this condition experienced the scrambled paragraphs in
the same order.  The word-scrambled experimental condition (word)
had participants listen to a temporally scrambled version of the story
where the words in the story occurred in a random order.  All
participants in the word condition experienced the scrambled words in
the same order.  Finally, in a rest experimental condition (rest),
participants lay in the scanner with no overt stimulus, with their
eyes open (blinking as needed).  This public dataset provided a convenient
means of testing our hypothesis that different levels of cognitive
processing and engagement are reflected in different orders of brain
activity dynamics.

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/decode_level}
  \caption{\textbf{Across-participant timepoint decoding accuracy varies with
       correlation order and cognitive engagement.}
     \textbf{a.~Decoding accuracy as a function of order: PCA.}
   ``Order'' ($x$-axis) refers to the maximum order of dynamic
    correlations that were available to the classifiers (see
    Feature weighting and testing).  The reported
    across-participant decoding accuracies are averaged over all
    kernel shapes and widths (see Identifying robust decoding
      results).  The $y$-values are displayed relative to chance
    accuracy (intact: $\frac{1}{300}$; paragraph: $\frac{1}{272}$;
    word: $\frac{1}{300}$; rest: $\frac{1}{400}$; these chance accuracies
    were subtracted from the observed accuracies to obtain the
    relative accuracies reported on the $y$-axis).  The error ribbons
    denote 95\% confidence intervals of the means across cross-validation folds
    (i.e., random assignments of participants to the training and test
    sets).  The colors denote the experimental condition.  Arrows
    denote sets of features that yielded reliably higher (upward
    facing) or lower (downward facing) decoding accuracy than the mean
    of all other features (via a two-tailed $t$-test, thresholded at
    $p < 0.05$).  Figure~\ref{fig:ttests} displays additional
    comparisons between the decoding accuracies achieved using
    different sets of neural features.  The circled values represent
    the maximum decoding accuracy within each experimental condition.
  \textbf{b.~Normalized timepoint decoding accuracy as a function of order:
       PCA.}  This panel displays the same results as Panel a, but here
    each curve has been normalized to have a maximum value of 1 and a
    minimum value of 0 (including the upper and lower bounds of the
    respective 95\% confidence intervals of the mean).  Panels a and b used PCA to
    project each high-dimensional pattern of dynamic correlations onto
    a lower-dimensional space. \textbf{c.~Timepoint decoding accuracy as a
       function of order: eigenvector centrality.} This panel is in the
    same format as Panel a, but here eigenvector centrality has been
    used to project the high-dimensional patterns of dynamic
    correlations onto a lower-dimensional space.
 \textbf{d.~Normalized timepoint decoding accuracy as a function of order:
       eigenvector centrality.}  This panel is in the same format as
    Panel b, but here eigenvector centrality has been used to project
    the high-dimensional patterns of dynamic correlations onto a
    lower-dimensional space. See
    Figures~\kernelshape~and~\kernelwidth~for decoding results broken
    down by kernel shape and width, respectively. Source data are provided as a Source Data file.}
  \label{fig:decoding}
\end{figure}
